---
title: "Chu analysis for Ouija"
output:
  html_document: default
  html_notebook: default
---

Load libraries and sceset:

```{r, load-libraries}
set.seed(942)

knitr::opts_chunk$set(fig.width = 5, fig.height = 4, cache = TRUE)

suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(viridis))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(stringr))
devtools::load_all("~/oxford/bnlfa/oui/ouija")



sce <- readRDS("../../data/scesets/chu-sce.rds")
print(sce)
```

# Analysis with Ouija

## Exploratory analysis

Get marker genes - first 9 are switch-like, final 2 are transient

```{r}
markers <- c("POU5F1", "NANOG", "SOX2", "EOMES", "CER1", "GATA4", "DKK4",
             "MYCT1", "PRDM1", "CDX1", "MSX2")
response_types <- c(rep("switch", 9), rep("transient", 2))
```

Create a marker sceset:

```{r}
sce_marker <- sce[markers, ]
print(sce_marker)
```

Have a look at PCA of top 500 most variable (scater default) and markers only:

```{r, fig.show = 'hold', fig.height = 3, fig.width = 4}
plotPCA(sce, colour_by = 'capture_time')
plotPCA(sce_marker, colour_by = 'capture_time')
```


## Fitting pseudotimes

We'll call Ouija (using VB for now), with fairly diffuse switching priors

```{r, message = FALSE, warning = FALSE, results = 'hide'}
oui <- ouija(sce_marker, response_types, iter = 3000,
             strength_sd = rep(5, 9))
```

Quick diagnostic to check things are mostly ok:

```{r}
plot(oui, what = 'diagnostic')
```

```{r}
saveRDS(oui, "../../data/oui_chu.rds")
```


Plot the gene output:

```{r, fig.width = 6, fig.height = 5}
plot_expression(oui)
```

Plot the switch times:

```{r}
plot_switch_times(oui)
switch_plot <- last_plot()
```

Plot the peak times

```{r, fig.width = 3, fig.height = 2}
plot_transient_times(oui)
```



## Comparison of discrete cell types

We can construct the confusion matrix to find discrete cell types:

```{r}
cmat <- confusion_matrix(oui)
cmo <- confusion_matrix_ordered(oui, cmat)
plot_confusion(oui, cmo)
```

And cluster it to find cell types

```{r}
sce_marker$cluster <- factor(cluster_confusion(cmat))
```

How many optimal clusters were found?

```{r}
length(unique(sce_marker$cluster))

```

And plot in pca space:

```{r}
plotPCA(sce_marker, colour_by = 'cluster')
```

How do these clusters compare to the capture time?

```{r}
print(table(sce_marker$capture_time, sce_marker$cluster))
```



# Comparison with other pseudotime methods

```{r}
gene_vars <- matrixStats::rowVars(exprs(sce))
cutoff <- sort(gene_vars, decreasing = TRUE)[501]
sce_highvar <- sce[gene_vars > cutoff, ]
```


```{r}
source("../../scripts/pseudotime_fitting_functions.R")
dpt_highvar <- fit_dpt_pseudotime(sce_highvar)
tscan_highvar <- fit_tscan_pseudotime(sce_highvar)
monocle_highvar <- fit_monocle_pseudotime(sce_highvar)


d <- data.frame(dpt_highvar_genes = dpt_highvar, monocle_highvar_genes = monocle_highvar,
           ouija_markers = map_pseudotime(oui), 
           tscan_highvar_genes = tscan_highvar)
GGally::ggpairs(d)
```

```{r}
ouija_pst <- map_pseudotime(oui)
df_cor <- data.frame(
  monocle = cor(d$monocle_highvar_genes, ouija_pst),
  dpt = cor(d$dpt_highvar_genes, ouija_pst),
  tscan = cor(d$tscan_highvar_genes, ouija_pst, use = "na")
)
```

Find the max cor amongs the other algos:

```{r}
cors <- c(
  cor(d$monocle_highvar_genes, d$dpt_highvar_genes),
  cor(d$dpt_highvar_genes, d$tscan_highvar_genes),
  cor(d$monocle_highvar_genes, d$tscan_highvar_genes)
)
df_cor$max_cor <- max(abs(cors))
```



```{r, cache = FALSE}
readr::write_csv(df_cor, "../../data/mvt_csv/chu.csv")

```

# Interpretable pseudotime

```{r}
## DELETE
oui <- readRDS("../../data/oui_chu.rds")
```
```{r}
reg_df <- regulation_df(oui)
sig_regs <- significant_regulation(reg_df)

reg_df_all <- inner_join(reg_df, sig_regs, by = "label")

ggplot(reg_df_all, aes(x = param_diffs, fill = signif)) +
  geom_histogram() + facet_wrap(~ label, scales = "free_x") +
  labs(x = "Difference in switch times", subtitle = "Posterior difference in switch times") +
  scale_fill_brewer(palette = "Set2")
```

Expression plot!

```{r}
expr_df <- as_data_frame(t(exprs(sce_marker))) %>% 
  mutate(pseudotime = rank(map_pseudotime(oui))) %>% 
  gather(gene, expression, -pseudotime)

transient_genes <- c("CDX1", "MSX2")

expr_df <- mutate(expr_df,
                  gene_type = case_when(
                    gene %in% transient_genes ~ "Transient",
                    !(gene %in% transient_genes) ~ "Switch-like"
                  ))
gene_order <- c(as.character(levels(switch_plot$data$Gene)), transient_genes)

expr_df$gene <- factor(expr_df$gene, levels = gene_order)

size10_axis <- theme(axis.title = element_text(size = 10))

ggplot(expr_df, aes(x = pseudotime, y = gene, fill = expression)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_viridis(name = "Expression") +
  theme(panel.background = element_blank(),
        legend.position = "bottom",
        strip.background = element_rect(colour="white", fill="white")) +
  facet_grid(gene_type ~ ., scales = "free_y", space = "free") +
  labs(y = "Gene", x = sprintf("Pseudotime order \u2192")) +
  scale_x_continuous(expand = c(0,0)) +
  size10_axis
  
expression_plot <- last_plot()
```


```{r}
switch_plot <- plot_switch_times(oui) + ylim(c(0,1)) + theme(legend.position = "bottom") +
  size10_axis
transient_plot <- plot_transient_times(oui) +
  theme(axis.title.y = element_text(margin = margin(r = 10))) +
  size10_axis

cowplot::plot_grid(switch_plot, transient_plot, ncol = 1, rel_heights = c(3.5,1),
                   labels = c("B", "C"), label_size = 12, vjust = c(1.5, 0))

middle_grid <- last_plot()
```

And reg df

```{r}
## What we're interested in - 
iint <- c(
  "EOMES - DKK4",
  "MYCT1 - PRDM1",
  "DKK4 - MYCT1",
  "DKK4 - CDX1"
)

reg_df_i <- filter(reg_df_all, label %in% iint)
reg_df_i$label <- factor(reg_df_i$label, levels = iint)

reg_df_i$signif <- str_to_title(as.character(reg_df_i$signif))
reg_df_i <- mutate(reg_df_i,
                   `Significant difference` = signif)


ggplot(reg_df_i, aes(x = param_diffs, fill = `Significant difference`)) +
  geom_histogram(bins = 50) + facet_wrap(~ label) +
  labs(x = "Posterior difference in regulation times", y = "Posterior samples") +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "top",
        legend.title = element_text(size = 10),
        strip.background = element_rect(colour="white", fill="white")) +
  size10_axis

diff_plot <- last_plot()
diff_plot_grid <- cowplot::plot_grid(NULL, diff_plot, NULL, rel_heights = c(0.3, 5, 0.7), ncol = 1)
```



Overall figure

```{r, cache = FALSE}
cowplot::plot_grid(expression_plot, middle_grid, diff_plot_grid,
                   labels = c("A", "", "D"), label_size = 12,
                   rel_widths = c(2.5,2,1.5), nrow = 1)
ggsave("../../figs/fig2.png", width = 12, height = 5)
```

