---
title: "Dulken analysis for Ouija"
output:
  html_document: default
  html_notebook: default
---


```{r, include = FALSE}
cor_to_tidy <- function(d, dataset) {
  dcor <- cor(d, use = "na")
  dcor_df <- as_data_frame(dcor) %>% 
    dplyr::mutate(row_alg = rownames(dcor)) %>% 
    gather(col_alg, correlation, -row_alg) %>% 
    mutate(dataset = dataset)
  lt <- as_data_frame(lower.tri(dcor))
  colnames(lt) <- colnames(dcor)
  lt_df <- as_data_frame(lt) %>% 
    dplyr::mutate(row_alg = rownames(dcor)) %>% 
    gather(col_alg, is_lower_tri, -row_alg)
  dcor_df <- inner_join(dcor_df, lt_df, by = c("row_alg", "col_alg"))
  dcor_df <- dplyr::filter(dcor_df, is_lower_tri)
  dcor_df
}
```

Load libraries and sceset:

```{r, load-libraries}
set.seed(942)

knitr::opts_chunk$set(fig.width = 5, fig.height = 4, cache = TRUE)

suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(tidyr))


devtools::load_all("~/oxford/bnlfa/oui/ouija")

sce <- readRDS("../../data/scesets/dulken-sce.rds")
print(sce)
```

# Analysis with Ouija

## Exploratory analysis

Get marker genes - first 9 are switch-like, final 2 are transient

```{r}
markers <- c("Id3", "Clu", "Rpl32", "Egfr",
             "Cdk4", "Cdk1", "Dlx2", "Dcx")
```

Create a marker sceset:

```{r}
sce_marker <- sce[markers, ]
print(sce_marker)
```

Have a look at PCA of top 500 most variable (scater default) and markers only:

```{r, fig.show = 'hold', fig.height = 3, fig.width = 4}
plotPCA(sce, colour_by = 'cell_type')
plotPCA(sce_marker, colour_by = 'cell_type')
```


## Fitting pseudotimes

We'll call Ouija (using VB for now), with fairly diffuse switching priors

```{r, message = FALSE, warning = FALSE, results = 'hide'}
oui <- ouija(sce_marker, iter = 3000,
             strength_sd = rep(5, 8))
```

Quick diagnostic to check things are mostly ok:

```{r}
plot(oui, what = 'diagnostic')
```

Plot the gene output:

```{r, fig.width = 6, fig.height = 5}
plot_expression(oui)
```

Plot the switch times:

```{r}
plot_switch_times(oui)
```




## Comparison of discrete cell types

We can construct the confusion matrix to find discrete cell types:

```{r}
cmat <- confusion_matrix(oui)
cmo <- confusion_matrix_ordered(oui, cmat)
plot_confusion(oui, cmo)
```

And cluster it to find cell types

```{r}
sce_marker$cluster <- factor(cluster_confusion(cmat))
```

How many optimal clusters were found?

```{r}
length(unique(sce_marker$cluster))
```

And plot in pca space:

```{r}
plotPCA(sce_marker, colour_by = 'cluster')
```

How do these clusters compare to the capture time?

```{r}
print(table(droplevels(sce_marker$cell_type), sce_marker$cluster))
```


# Comparison with other pseudotime methods

```{r}
gene_vars <- matrixStats::rowVars(exprs(sce))
cutoff <- sort(gene_vars, decreasing = TRUE)[501]
sce_highvar <- sce[gene_vars > cutoff, ]
```

```{r}
source("../../scripts/pseudotime_fitting_functions.R")
dpt_highvar <- fit_dpt_pseudotime(sce_highvar)
tscan_highvar <- fit_tscan_pseudotime(sce_highvar)
monocle_highvar <- fit_monocle_pseudotime(sce_highvar)

d <- data.frame(dpt_highvar_genes = dpt_highvar, monocle_highvar_genes = monocle_highvar,
           ouija_markers = map_pseudotime(oui), 
           tscan_highvar_genes = tscan_highvar)
GGally::ggpairs(d)

```

```{r, cache = FALSE}
write_csv(cor_to_tidy(d, "dulken"), "../../data/cor_comp/dulken.csv")
```


```{r}
ouija_pst <- map_pseudotime(oui)
df_cor <- data.frame(
  monocle = cor(d$monocle_highvar_genes, ouija_pst),
  dpt = cor(d$dpt_highvar_genes, ouija_pst),
  tscan = cor(d$tscan_highvar_genes, ouija_pst, use = "na")
)
```


```{r}
cors <- c(
  cor(d$monocle_highvar_genes, d$dpt_highvar_genes),
  cor(d$dpt_highvar_genes, d$tscan_highvar_genes),
  cor(d$monocle_highvar_genes, d$tscan_highvar_genes)
)
df_cor$max_cor <- max(abs(cors))
```


```{r, cache = FALSE}
readr::write_csv(df_cor, "../../data/mvt_csv/dulken.csv")

```
