---
title: "Li analysis for Ouija"
output:
  html_document: default
  html_notebook: default
---

Load libraries and sceset:

```{r, load-libraries}

theme_bw_nolines <- function() {
  theme_bw() +
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank())
}

set.seed(942)

knitr::opts_chunk$set(fig.width = 5, fig.height = 4, cache = TRUE)

suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))

suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(rstan))

devtools::load_all("~/oxford/bnlfa/oui/ouija")

sce <- readRDS("../../data/scesets/li-sce.rds")
print(sce)
```

# Analysis with Ouija

## Exploratory analysis

Get marker genes - first 9 are switch-like, final 2 are transient

```{r}
markers <- c("Mef2c", "Foxa2", "Col10a1", "Comp", "Dlx3", "Id3",
             "Pik3r2", "Spry2", "Sox9", "Insig1",
             "Scd1", "Scd2")
stopifnot(all(markers %in% featureNames(sce)))
response_types <- c(rep("switch", 10), rep("transient", 2))
```

Create a marker sceset:

```{r}
sce_marker <- sce[markers, ]
print(sce_marker)
```

Have a look at PCA of top 500 most variable (scater default) and markers only. We don't have any additional phenotype data, but the authors note that pseudotime tracks Col10a expression:

```{r, fig.show = 'hold', fig.height = 3, fig.width = 4}
plotPCA(sce, colour_by = 'Col10a1')
plotPCA(sce_marker, colour_by = 'Col10a1')
```


## Fitting pseudotimes

We'll call Ouija (using VB for now), with fairly diffuse switching priors

```{r, message = FALSE, warning = FALSE, results = 'hide'}
oui_file <-  "../../data/ouija_fits/li.rds"
if(!file.exists(oui_file)) {
  oui <- ouija(sce_marker, response_types, iter = 3000, 
               strength_sd = rep(5, 10))
  saveRDS(oui, oui_file)
} else {
  oui <- readRDS(oui_file)
}
```



Quick diagnostic to check things are mostly ok:

```{r}
plot(oui, what = 'diagnostic')
```

Plot the gene output:

```{r, fig.width = 6, fig.height = 5}
plot_expression(oui)
```

Plot the switch times:

```{r}
plot_switch_times(oui)
```




## Comparison of discrete cell types

We can construct the confusion matrix to find discrete cell types:

```{r}
cmat <- confusion_matrix(oui)
cmo <- confusion_matrix_ordered(oui, cmat)
plot_confusion(oui, cmo)
```

And cluster it to find cell types

```{r}
sce_marker$cluster <- factor(cluster_confusion(cmat))
```

How many optimal clusters were found?

```{r}
length(unique(sce_marker$cluster))
```

And plot in pca space:

```{r}
plotPCA(sce_marker, colour_by = 'cluster')
```




# Comparison with other pseudotime methods

```{r}
gene_vars <- matrixStats::rowVars(exprs(sce))
cutoff <- sort(gene_vars, decreasing = TRUE)[501]
sce_highvar <- sce[gene_vars > cutoff, ]
```

```{r}
source("../../scripts/pseudotime_fitting_functions.R")
dpt_highvar <- fit_dpt_pseudotime(sce_highvar)
tscan_highvar <- fit_tscan_pseudotime(sce_highvar)
monocle_highvar <- fit_monocle_pseudotime(sce_highvar)

dpt_highvar[which.min(dpt_highvar)] <- NA

d <- data.frame(dpt_highvar_genes = dpt_highvar, monocle_highvar_genes = monocle_highvar,
           ouija_markers = colMeans(rstan::extract(oui$fit, "t")$t), 
           tscan_highvar_genes = tscan_highvar)
```

```{r}
ouija_pst <- colMeans(rstan::extract(oui$fit, "t")$t)
df_cor <- data.frame(
  monocle = cor(d$monocle_highvar_genes, ouija_pst),
  dpt = cor(d$dpt_highvar_genes, ouija_pst, use = "na"),
  tscan = cor(d$tscan_highvar_genes, ouija_pst, use = "na")
)

```


```{r}
cors <- c(
  cor(d$monocle_highvar_genes, d$dpt_highvar_genes, use = "na"),
  cor(d$dpt_highvar_genes, d$tscan_highvar_genes, use = "na"),
  cor(d$monocle_highvar_genes, d$tscan_highvar_genes, use = "na")
)
df_cor$max_cor <- max(abs(cors))
```

```{r}
cor_mat <- cor(d, use = "na")
cor_mat <- abs(cor_mat)
diag(cor_mat) <- "NA"

cor_df <- as_data_frame(cor_mat) %>% 
  mutate(alg1 = rownames(cor_mat)) %>% 
  gather(alg2, correlation, -alg1) 
cor_df$correlation <- as.numeric(cor_df$correlation)

tidy_algorithm_str <- function(s) {
  sapply(stringr::str_split(s, "_"), `[`, 1) %>% 
    stringr::str_to_title() 
}

uppercase_dpt_tscan <- function(s) {
  if(s == "Dpt" ||  s == "Tscan") return(stringr::str_to_upper(s))
  s
}

cor_df$alg1_tidy <- tidy_algorithm_str(cor_df$alg1)
cor_df$alg2_tidy <- tidy_algorithm_str(cor_df$alg2)

cor_df$alg1_tidy <- sapply(cor_df$alg1_tidy, uppercase_dpt_tscan)
cor_df$alg2_tidy <- sapply(cor_df$alg2_tidy, uppercase_dpt_tscan)

ggplot(cor_df, aes(x = alg1_tidy, y = alg2_tidy, fill = correlation)) +
  geom_raster() +
  viridis::scale_fill_viridis(name = "Absolute\ncorrelation") +
  labs(x = "Algorithm", y = "Algorithm") +
    theme(panel.background = element_blank(),
          axis.text = element_text(size = 10),
          legend.title = element_text(size = 10))

cor_plot <- last_plot()
```



```{r, cache = FALSE}
readr::write_csv(df_cor, "../../data/mvt_csv/li.csv")
```


```{r}
d2 <- d
d2$original_waterfall_pseudotime <- sce$wpst

col10a1 <- exprs(sce)["Col10a1",]

for(i in 1:(ncol(d2) - 1)) {
  d_i <- d2[,i]
  if(cor(d_i, col10a1, use = "na") < 0)
    d2[,i] <- max(d_i) - d_i
}

df_gex <- as_data_frame(t(exprs(sce_marker))) %>% 
  bind_cols(d2)
df_gex_tidy1 <- gather(df_gex, gene, expression, -(dpt_highvar_genes:tscan_highvar_genes))
df_gex_tidy <- gather(df_gex_tidy1, algorithm, pseudotime, -gene, -expression)

ggplot(df_gex_tidy, aes(x = pseudotime, y = expression)) +
  geom_point() +
  geom_smooth() +
  facet_grid(gene ~ algorithm, scales = "free")

ggsave("../../figs/supp_li.png", width = 11, height = 11)
```

Example genes

```{r}
gene_df <- frame_data(
  ~ gene, ~ original_behaviour, ~ designation,
  "Mef2c", "Upregulation", "Correct",
  "Pik3r2", "Late upregulation", "Correct",
  # "Comp", "Upregulation", "Incorrect",
  "Scd1", "Transient", "Incorrect"
)

expr_df <- as_data_frame(t(exprs(sce_marker)[gene_df$gene, ])) %>% 
  mutate(ouija_pseudotime = d$ouija_markers) %>% 
  gather(gene, expression, -ouija_pseudotime)

expr_df <- inner_join(expr_df, gene_df, by = "gene")
expr_df$gene <- factor(expr_df$gene, levels = gene_df$gene)

cols <- c("#4575b4", "#d73027")

plot_expr <- function(df, co) {
  ggplot(df, aes(x = ouija_pseudotime, y = expression)) +
    geom_point(color = co) +
    facet_wrap(~ gene, scales = "free_y", ncol = 1) +
    stat_smooth(color = 'black') +
    theme_bw_nolines() +
    theme(strip.background = element_rect(colour="white", fill="white"),
          axis.title = element_text(size = 10)) +
    labs(x = "Ouija pseudotime", y = "Expression")
}

correct_genes <- filter(expr_df, designation == "Correct") %>% 
  plot_expr(cols[1])
incorrect_genes <- filter(expr_df, designation == "Incorrect") %>% 
  plot_expr(cols[2])
```

```{r, cache = FALSE}
saveRDS(cor_plot, "../../figs/li_cor.rds")
saveRDS(correct_genes, "../../figs/li_correct_genes.rds")
saveRDS(incorrect_genes, "../../figs/li_incorrect_genes.rds")
```

