---
title: "Trapnell analysis for Ouija"
output:
  html_document: default
  html_notebook: default
---

Load libraries and sceset:

```{r, load-libraries}
set.seed(942)

knitr::opts_chunk$set(fig.width = 5, fig.height = 4, cache = TRUE)

suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(ggplot2))
devtools::load_all("~/oxford/bnlfa/oui/ouija")


sce_list <- readRDS("../../data/scesets_with_marker_pseudotime.rds")
sce <- sce_list[['shin']]
print(sce)
```

# Analysis with Ouija

## Exploratory analysis



Create a marker sceset:

```{r}
sce_marker <- sce[fData(sce)$is_marker, ]
print(sce_marker)
```

Have a look at PCA of top 500 most variable (scater default) and markers only. We don't have any additional phenotype data, but the authors note that pseudotime tracks Col10a expression:

```{r, fig.show = 'hold', fig.height = 3, fig.width = 4}
plotPCA(sce)
plotPCA(sce_marker)
```


## Fitting pseudotimes

We'll call Ouija (using VB for now), with fairly diffuse switching priors

```{r, message = FALSE, warning = FALSE, results = 'hide'}
oui <- ouija(sce_marker, iter = 3000, 
             strength_sd = rep(5, 6))
```

Quick diagnostic to check things are mostly ok:

```{r}
plot(oui, what = 'diagnostic')
```



Plot the gene output:

```{r, fig.width = 6, fig.height = 5}
plot_expression(oui)
```

Plot the switch times:

```{r}
plot_switch_times(oui)
```




## Comparison of discrete cell types

We can construct the confusion matrix to find discrete cell types:

```{r}
cmat <- confusion_matrix(oui)
cmo <- confusion_matrix_ordered(oui, cmat)
plot_confusion(oui, cmo)
```

And cluster it to find cell types

```{r}
sce_marker$cluster <- factor(cluster_confusion(cmat))
```

How many optimal clusters were found?

```{r}
length(unique(sce_marker$cluster))
```

And plot in pca space:

```{r}
plotPCA(sce_marker, colour_by = 'cluster')
```



# Comparison with other pseudotime methods

```{r}
gene_vars <- matrixStats::rowVars(exprs(sce))
cutoff <- sort(gene_vars, decreasing = TRUE)[501]
sce_highvar <- sce[gene_vars > cutoff, ]
```

```{r}

source("../../scripts/pseudotime_fitting_functions.R")
dpt_highvar <- fit_dpt_pseudotime(sce_highvar)
tscan_highvar <- fit_tscan_pseudotime(sce_highvar)
monocle_highvar <- fit_monocle_pseudotime(sce_highvar, ncenter = 50)

d <- data.frame(dpt_highvar_genes = dpt_highvar, monocle_highvar_genes = monocle_highvar,
           ouija_markers = map_pseudotime(oui), 
           tscan_highvar_genes = tscan_highvar)
GGally::ggpairs(d)
```

```{r}
ouija_pst <- map_pseudotime(oui)
df_cor <- data.frame(
  monocle = cor(d$monocle_highvar_genes, ouija_pst),
  dpt = cor(d$dpt_highvar_genes, ouija_pst),
  tscan = cor(d$tscan_highvar_genes, ouija_pst, use = "na")
)

```


```{r}
cors <- c(
  cor(d$monocle_highvar_genes, d$dpt_highvar_genes),
  cor(d$dpt_highvar_genes, d$tscan_highvar_genes, use = "na"),
  cor(d$monocle_highvar_genes, d$tscan_highvar_genes, use = "na")
)
df_cor$max_cor <- max(abs(cors))
```


```{r}
readr::write_csv(df_cor, "../../data/mvt_csv/shin.csv")

```
