---
title: "Trapnell analysis for Ouija"
output:
  html_document: default
  html_notebook: default
---

Load libraries and sceset:

```{r, load-libraries}
set.seed(942)

knitr::opts_chunk$set(fig.width = 5, fig.height = 4, cache = TRUE)

suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(gplots))
suppressPackageStartupMessages(library(viridis))
devtools::load_all("~/oxford/bnlfa/oui/ouija")

source("../../scripts/pseudotime_fitting_functions.R")

sce_list <- readRDS("../../data/scesets_with_marker_pseudotime.rds")
sce <- sce_list[['trapnell']]
print(sce)
```

# Analysis with Ouija

## Exploratory analysis



Create a marker sceset:

```{r}
sce_marker <- sce[fData(sce)$is_marker, ]
print(sce_marker)
```

Have a look at PCA of top 500 most variable (scater default) and markers only. We don't have any additional phenotype data, but the authors note that pseudotime tracks Col10a expression:

```{r, fig.show = 'hold', fig.height = 3, fig.width = 4}
plotPCA(sce, colour_by = 'Hours')
plotPCA(sce_marker, colour_by = 'Hours')
```


## Fitting pseudotimes

We'll call Ouija (using VB for now), with fairly diffuse switching priors

```{r, message = FALSE, warning = FALSE, results = 'hide'}
oui <- ouija(sce_marker, iter = 3000, 
             strength_sd = rep(5, 5))
```

Quick diagnostic to check things are mostly ok:

```{r}
plot(oui, what = 'diagnostic')
```




Plot the gene output:

```{r, fig.width = 6, fig.height = 5}
plot_expression(oui)
```

Plot the switch times:

```{r}
plot_switch_times(oui)
```




## Comparison of discrete cell types

We can construct the confusion matrix to find discrete cell types:

```{r}
cmat <- confusion_matrix(oui)
cmo <- confusion_matrix_ordered(oui, cmat)
plot_confusion(oui, cmo)
```

And cluster it to find cell types

```{r}
sce_marker$cluster <- factor(cluster_confusion(cmat))
```

How many optimal clusters were found?

```{r}
length(unique(sce_marker$cluster))
```

And plot in pca space:

```{r}
plotPCA(sce_marker, colour_by = 'cluster')
```


Comparison with other pseudotime methods

```{r}
gene_vars <- matrixStats::rowVars(exprs(sce))
cutoff <- sort(gene_vars, decreasing = TRUE)[501]
sce_highvar <- sce[gene_vars > cutoff, ]
```

```{r}
dpt_highvar <- fit_dpt_pseudotime(sce_highvar)
tscan_highvar <- fit_tscan_pseudotime(sce_highvar)

d <- data.frame(original_monocle = sce$Pseudotime, dpt_highvar_genes = dpt_highvar,
           ouija_markers = map_pseudotime(oui), 
           tscan_highvar_genes = tscan_highvar)
GGally::ggpairs(d)
```

```{r}
ouija_pst <- map_pseudotime(oui)
df_cor <- data.frame(
  monocle = cor(d$original_monocle, ouija_pst),
  dpt = cor(d$dpt_highvar_genes, ouija_pst),
  tscan = cor(d$tscan_highvar_genes, ouija_pst)
)
```

Find the max cor amongs the other algos:

```{r}
cors <- c(
  cor(d$original_monocle, d$dpt_highvar_genes),
  cor(d$dpt_highvar_genes, d$tscan_highvar_genes),
  cor(d$original_monocle, d$tscan_highvar_genes)
)
df_cor$max_cor <- max(abs(cors))
```



```{r, cache = FALSE}
readr::write_csv(df_cor, "../../data/mvt_csv/trapnell.csv")
```

# Comparison of gene fits

Construct expression df

```{r}
example_genes <- c("MYOG", "ID1")
example_inds <- match(example_genes, fData(sce_marker)$gene_short_name)
expr <- oui$Y[, example_inds]
colnames(expr) <- example_genes
```


First sanitize d

```{r}
scale_to_unif <- function(x) (x - min(x)) / (max(x) - min(x)) # scale to lie in [0, 1]
d2 <- as_data_frame(d)
d2 <- apply(d2, 2, scale_to_unif) 

## Correlations with MYOG - flip if negative
cors_with_myog <- apply(d2, 2, cor, expr[,1])
d2[, cors_with_myog < 0] <- 1 - d2[, cors_with_myog < 0]
d2 <- as_data_frame(d2)
```



```{r}

expr_df <- cbind(expr, d2)

map_from <- colnames(d2)
map_to <- c("Monocle (original)", "DPT", "Ouija (markers)", "TSCAN")

expr_df_tidy1 <- gather(expr_df, algorithm, pseudotime, -MYOG, -ID1)
expr_df_tidy2 <- gather(expr_df_tidy1, gene, expression, -algorithm, -pseudotime)
expr_df_tidy2$algorithm <- plyr::mapvalues(expr_df_tidy2$algorithm, from = map_from, to = map_to)
expr_df_tidy2 <- as_data_frame(expr_df_tidy2)

filter(expr_df_tidy2, !grepl("Ouija", algorithm)) %>% 
  ggplot(aes(x = pseudotime, y = expression)) +
    geom_point(shape = 21, color = 'black', fill = 'grey70') +
    facet_grid(algorithm ~ gene) +
    labs(x = "Pseudotime", y = "Normalised expression") +
    stat_smooth(se = F, color = 'darkred') +
  scale_y_continuous() +
  theme(strip.background = element_rect(colour="white", fill="white")) +
  theme(axis.title = element_text(size = 9))

algorithm_plot <- last_plot()
```

Now do Ouija separately

```{r}
pred_expr <- predicted_expression(oui)[, example_inds]
colnames(pred_expr) <- example_genes
pred_expr <-  as_data_frame(pred_expr) %>% 
  mutate(pseudotime = d2[['ouija_markers']], algorithm = "Ouija (markers)") %>% 
  gather(gene, expression, -pseudotime, -algorithm)

pred_expr <- arrange(pred_expr, pseudotime)

filter(expr_df_tidy2, grepl("Ouija", algorithm)) %>% 
  ggplot(aes(x = pseudotime, y = expression)) +
    geom_point(shape = 21, color = 'black', fill = 'grey70') +
    facet_grid(algorithm ~ gene) +
    labs(x = "Pseudotime", y = "Normalised expression") +
  theme(strip.background = element_rect(colour="white", fill="white")) +
  theme(axis.title = element_text(size = 9))

base_plot <- last_plot()

sample_tidy <- function() {
  expr_samples <- sample_predicted_expression(oui)[, example_inds]
  colnames(expr_samples) <- example_genes
  expr_samples <-  as_data_frame(expr_samples) %>% 
  mutate(pseudotime = d2[['ouija_markers']], algorithm = "Ouija (markers)") %>% 
  gather(gene, expression, -pseudotime, -algorithm)
  expr_samples <- arrange(expr_samples, pseudotime)
  expr_samples
}

for(i in 1:40) base_plot <- base_plot + geom_line(data = sample_tidy(), alpha = 0.3)
plt <- base_plot + geom_line(data = pred_expr, color = 'darkred', size = 1.2)

```

Overall plot:

```{r}
cowplot::plot_grid(plt, algorithm_plot, labels = c("C", "D"),
                   rel_heights = c(1,2.2), ncol = 1, label_size = 12)
example_plot <- last_plot()
```

```{r, cache = FALSE}
saveRDS(last_plot(), "../../figs/trapnell_example_genes.rds")

```


Correlation figure


```{r}
ecor <- cor(t(exprs(sce_highvar)))
png(file.path(tempdir(), "heatmap.png"))
hmap <- heatmap.2(ecor, trace = 'none', col = "viridis", dendrogram = "none")
dev.off()
```

Now do it nicely in ggplot:


```{r}
ecor_reordered <- ecor[hmap$rowInd, hmap$colInd]

## Where are our markers?
marker_df <- data_frame(markers = fData(sce_highvar)[['gene_short_name']][fData(sce_highvar)[['is_marker']]],
                        index = which(fData(sce_highvar)[['is_marker']]))
match(marker_df$index, hmap$rowInd)
```

```{r, cache = FALSE}
diag(ecor_reordered) <- NA
er <- as_data_frame(ecor_reordered)
names(er) <- seq_len(ncol(er))
er$row <- seq_len(nrow(er))
er_tidy <- gather(er, column, correlation, -row)
er_tidy$column <- as.integer(er_tidy$column)

pal <- viridis_pal()(9)

ggplot(er_tidy, aes(x = row, y = column, fill = correlation)) +
  geom_raster() + #scale_fill_gradient2(low = pal[1], mid = pal[5], high = pal[9])
  scale_fill_viridis(option = "C", name = "Correlation") +
  theme_void() #+
  # theme(legend.position = 'top')
saveRDS(last_plot(), "../../figs/trapnell_correlation.rds")
```

```{r}
pca <- prcomp(t(exprs(sce_highvar)))
s <- summary(pca)$importance
s_top <- s[2,1:6]

df_pca <- data_frame(pcs = names(s_top), pct_expl = 100 * s_top)


ggplot(df_pca, aes(x = pcs, y = pct_expl)) + 
  geom_bar(stat = "identity") +
  ylab("% Variance explained") + 
  xlab("Principal component") +
  theme(axis.title = element_text(size = 10))

scree_plot <- last_plot()

```

```{r}
pc1 <- pca$x[,1]
pc2 <- pca$x[,2]

pc1_cors <- apply(exprs(sce), 1, cor, pc1)
pc2_cors <- apply(exprs(sce), 1, cor, pc2)

pc_cor_df <- data_frame(PC1 = pc1_cors, PC2 = pc2_cors) %>% 
  gather(PC, correlation)

filter(pc_cor_df, abs(correlation) > 0) %>% 
  ggplot(aes(x = (correlation))) +
  geom_histogram() +
  facet_wrap(~ PC) +
  theme(axis.title = element_text(size = 10),
        strip.background = element_rect(colour="white", fill="white")) +
  labs(y = "Number of genes", x = "Correlation to principal component")

pc_cor_plot <- last_plot()
```

