---
title: "Zhou analysis for Ouija"
output:
  html_document: default
  html_notebook: default
---


```{r, include = FALSE}
cor_to_tidy <- function(d, dataset) {
  dcor <- cor(d, use = "na")
  dcor_df <- as_data_frame(dcor) %>% 
    dplyr::mutate(row_alg = rownames(dcor)) %>% 
    gather(col_alg, correlation, -row_alg) %>% 
    mutate(dataset = dataset)
  lt <- as_data_frame(lower.tri(dcor))
  colnames(lt) <- colnames(dcor)
  lt_df <- as_data_frame(lt) %>% 
    dplyr::mutate(row_alg = rownames(dcor)) %>% 
    gather(col_alg, is_lower_tri, -row_alg)
  dcor_df <- inner_join(dcor_df, lt_df, by = c("row_alg", "col_alg"))
  dcor_df <- dplyr::filter(dcor_df, is_lower_tri)
  dcor_df
}
```

Load libraries and sceset:

```{r, load-libraries}
set.seed(942)

knitr::opts_chunk$set(fig.width = 5, fig.height = 4, cache = TRUE)

suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(viridis))
suppressPackageStartupMessages(library(gridExtra))
devtools::load_all("~/oxford/bnlfa/oui/ouija")


sce_list <- readRDS("../../data/scesets_with_marker_pseudotime.rds")
sce <- sce_list[['hsc']]
print(sce)
```

# Analysis with Ouija

## Exploratory analysis

Create a marker sceset:

```{r}
sce_marker <- sce[fData(sce)$is_marker, ]
print(sce_marker)
```

Have a look at PCA of top 500 most variable (scater default) and markers only. We don't have any additional phenotype data, but the authors note that pseudotime tracks Col10a expression:

```{r, fig.show = 'hold', fig.height = 3, fig.width = 4}
plotPCA(sce, colour_by = 'cell_type')
plotPCA(sce_marker, colour_by = 'cell_type')
```

```{r}
sce$`Cell type` <- sce$cell_type
plotPCA(sce, colour_by = 'Cell type', ntop = 1000)
ggsave("../../figs/supp_zhou_pca.png", width = 5, height = 4)
```


## Fitting pseudotimes

We'll call Ouija (using VB for now), with fairly diffuse switching priors

```{r, message = FALSE, warning = FALSE, results = 'hide'}
oui <- ouija(sce_marker, iter = 3000, 
             strength_sd = rep(5, 6))
```

Quick diagnostic to check things are mostly ok:

```{r}
plot(oui, what = 'diagnostic')
```



Plot the gene output:

```{r, fig.width = 6, fig.height = 5}
plot_expression(oui)
```

Plot the switch times:

```{r}
plot_switch_times(oui)
```




## Comparison of discrete cell types

We can construct the confusion matrix to find discrete cell types:

```{r}
cmat <- confusion_matrix(oui)
cmo <- confusion_matrix_ordered(oui, cmat)
plot_confusion(oui, cmo)

cplot <- last_plot()
```

And cluster it to find cell types

```{r}
sce_marker$cluster <- factor(cluster_confusion(cmat))
```

How many optimal clusters were found?

```{r}
length(unique(sce_marker$cluster))
```

And plot in pca space:

```{r}
plotPCA(sce_marker, colour_by = 'cluster')
```

```{r}
tbl <- table(sce_marker$cluster, sce_marker$cell_type) 
print(tbl)
```

Turn this into a plot

```{r}
tdf <- as_data_frame(t(t(tbl) / colSums(tbl))) %>% 
  rename(Ouija_cluster = Var1, cell_type = Var2)

ggplot(tdf, aes(x = cell_type, y = Ouija_cluster, fill = n)) + 
  geom_raster() + 
  # scale_fill_gradient(name = "Proportion of cells assigned") +
  scale_fill_viridis(name = "Proportion of cells assigned") +
  theme(legend.position = "top") +
  labs(x = "Cell type", y = "Ouija cluster")

  
```

```{r}
sce_marker$map_pseudotime <- map_pseudotime(oui)
plot(rank(sce_marker$map_pseudotime), sce_marker$cluster)
decision_boundaries <- c(19, 79)

```



# Dynamics


```{r}
reg_df <- regulation_df(oui)
sig_regs <- significant_regulation(reg_df)

reg_df_all <- inner_join(reg_df, sig_regs, by = "label")

ggplot(reg_df_all, aes(x = param_diffs, fill = signif)) +
  geom_histogram() + facet_wrap(~ label) +
  labs(x = "Difference in switch times", subtitle = "Posterior difference in switch times") +
  scale_fill_brewer(palette = "Set2")
```


# Comparison with other pseudotime methods

```{r}
gene_vars <- matrixStats::rowVars(exprs(sce))
cutoff <- sort(gene_vars, decreasing = TRUE)[501]
sce_highvar <- sce[gene_vars > cutoff, ]
```

```{r}
source("../../scripts/pseudotime_fitting_functions.R")
dpt_highvar <- fit_dpt_pseudotime(sce_highvar)
tscan_highvar <- fit_tscan_pseudotime(sce_highvar)
monocle_highvar <- fit_monocle_pseudotime(sce_highvar)

d <- data.frame(dpt_highvar_genes = dpt_highvar, monocle_highvar_genes = monocle_highvar,
           ouija_markers = map_pseudotime(oui), 
           tscan_highvar_genes = tscan_highvar)
GGally::ggpairs(d)

```

```{r, cache = FALSE}
write_csv(cor_to_tidy(d, "zhou"), "../../data/cor_comp/zhou.csv")
```


```{r}
ouija_pst <- map_pseudotime(oui)
df_cor <- data.frame(
  monocle = cor(d$monocle_highvar_genes, ouija_pst),
  dpt = cor(d$dpt_highvar_genes, ouija_pst),
  tscan = cor(d$tscan_highvar_genes, ouija_pst, use = "na")
)
```


```{r}
cors <- c(
  cor(d$monocle_highvar_genes, d$dpt_highvar_genes),
  cor(d$dpt_highvar_genes, d$tscan_highvar_genes, use = "na"),
  cor(d$monocle_highvar_genes, d$tscan_highvar_genes, use = "na")
)
df_cor$max_cor <- max(abs(cors))
```

```{r}
decision_boundaries_pst <- sort(map_pseudotime(oui))[decision_boundaries]

expr_df <- as_data_frame(oui$Y) %>% 
  mutate(ouija_pseudotime = map_pseudotime(oui)) %>% 
  gather(gene, expression, -ouija_pseudotime)


mu_df <- as_data_frame(predicted_expression(oui)) %>% 
  mutate(ouija_pseudotime = map_pseudotime(oui)) %>% 
  gather(gene, expression, -ouija_pseudotime) %>% 
  arrange(ouija_pseudotime)

bg_df <- data_frame(Ouija_cell_type = as.factor(1:3),
                 xlower = c(0, decision_boundaries_pst),
                 xupper = c(decision_boundaries_pst, 1),
                 ylower = -Inf, 
                 yupper = Inf)

expression_plot <- ggplot(expr_df) +
  geom_rect(data = bg_df, aes(xmin = xlower, xmax = xupper, 
                               ymin = ylower, ymax = yupper, fill = Ouija_cell_type),
                alpha = 0.4) +
  geom_point(aes(x = ouija_pseudotime, y = expression), alpha = 0.65, color = 'black') + 
  facet_wrap(~ gene, ncol = 2, 
             strip.position = 'top') +
  geom_line(data = mu_df, aes(x = ouija_pseudotime, y = expression), size = 1.2, alpha = 0.7,
            color = 'black') +
  scale_color_brewer(palette = "Set2", name = "Cell type") +
  theme(legend.position = "top",
        strip.background = element_rect(fill="white")) +
  ylab("Normalised log expression") +
  xlab("Ouija pseudotime") +
  scale_y_continuous(breaks = c(0, 1, 2, 3), expand = c(0, 0))  +
  scale_x_continuous(expand = c(0, 0), breaks = c(0, 0.2, 0.4, 0.6, 0.8)) +
    scale_fill_brewer(name = "Ouija cell type", palette = "Set1") +
    theme(legend.box.margin = margin(b = -10))
```

Finally sort the table

```{r}
colnames(tbl) <- c("EC", "T1-preHSC", "T2-preHSC", "E12-HSC", "E14-HSC")
rownames(tbl) <- paste("Ouija cell type", 1:3)
grid.table(tbl)
```

Overall figure:

```{r, cache = FALSE}
left_grid <- cowplot::plot_grid(cplot, NULL, ncol = 1, rel_heights = c(2,1),
                                labels = c("A", "B"), label_size = 12)
cplot <- cplot + labs(x = "Pseudotime order", y = "Pseudotime order")
cowplot::plot_grid(left_grid, expression_plot, ncol = 2, labels = c("", "C"), label_size = 12)
ggsave("../../figs/fig3.png", width = 10, height = 6)
```



```{r, cache = FALSE}
readr::write_csv(df_cor, "../../data/mvt_csv/zhou.csv")
```
